Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-05-12 doc/temp-sentiment.tex

\documentclass{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[a4paper, margin=1.25in]{geometry}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage[
  style=numeric,
  citestyle=numeric-comp,
  backend=biber,
  doi=true,
  natbib=true,
  sorting=none
]{biblatex}
\usepackage{scalerel,xparse}

\addbibresource{TextDataClimateShocks.bib}

@@

\begin{document}
%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures}

@@

\maketitle

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in 250 million geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effects is much stronger in poor and black neighborhoods.

\end{abstract}

\section{Main}
%Overview
Climate change is impacting many aspects of human well-being with these impacts projected to worsen in the coming years \cite{pachauri2014climate}.  A large body of research has assessed the ongoing impacts on climate change on outcomes like agriculture, economic growth, and physical health (NEED CITATIONS). Academic and medical researchers have highlighted the relative paucity of empirical studies into climate impacts on mental health at multiple scales and recently called for more \cite{Berry2018Apr, hayes_climate_2018}.  Researchers have begun to answer these calls, finding associations between deteriorating environmental conditions and a sense of "ecological grief" from their experienced loss \cite{Cunsolo2018Apr}, disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, as well as a variety of mental health impacts associated with warmer temperatures \cite{baylis_weather_2018, Mullins2019Dec} (PALINKAS and WONG 2020).  MAYBE A SENTENCE HERE HIGHLIGHTING THE GAP IN KNOWLEDGE ABOUT DISPARITIES IN MENTAL HEALTH IMPACTS, VULNERABLE POPULATIONS, issues of scale for signal to noise, etc.

%Heat and mental health
%https://www.pnas.org/content/115/43/10953
Higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \citep{Burke2018Aug, Mullins2019Dec, Dixon2007May} with others demonstrating similar relationships in other locations around the world \citep{Qi2014Dec, Page2007Aug, Likhvar2011Jan}.  Higher temperatures have also been correlated with increases in hospitalization events related to both general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} and specific conditions like bipolar disorder and schizophrenia \citep{Lee2007Jan, Shapira2004Feb, Sung2013Feb, Gupta1992Jun, Hansen2008Oct}.

@@

Additionally, recent studies have found that nighttime temperatures affects sleep quality \citep{Obradovich2017May, Mullins2019Dec}.

% The pathways
Higher temperature associated with climate change could influence mental health directly by exposing people to trauma. It could also indirectly influence mental health by affecting (1) physical health and (2)community well-being \citep{RN1314}. For the direct influence pathway, extreme heat events or increasing temperatures have been associated with the increase in aggression,higher suicide rates and other hospital admissions \citep{RN1314,RN1316,RN1317,RN1318}. Heat exposures in working environments could also reduce people's capacity to deal with physical and mental task, and increase the risk of accidents, because of the excessive core body temperature and dehydration \citep{RN1319,RN1320,RN1321,RN1322}. The loss of work capacity would in turn result in loss of income, while the low income could also cause mental health problems \citep{Katz1997}.  As for the indirect pathway, first, because of the reciprocal causation relationship between physical health problems and mental health problems \citep{RN1323,RN1324} heat exposures associated with climate change along with other climate events and indirect health risks threatening physical health will directly influence mental health \citep{RN1325,RN1326}. Second, disordered temperature associated with climate change may destroy the economic in agricultural-production-dependent communities. For example, extreme heat reduces the work capacity of laborers in farm fields \citep{RN1320}, which further destroys agricultural-supported industries in local area\citep{RN1327}. The following economic pressures would undermine social capital and then lead to mental health problems.

@@

\section{Results}

\subsection{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  Controlling for a variety of fixed effects, we found that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree Celsius (a dry bulb temperature typically around 12\textdegree C/54\textdegree F).  The greatest declines in sentiment are observed between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We ran models using multiple sentiment indicators, and found similar results no matter which sentiment metric was used (See Supplement).

@@

\subsection{Comparison With Other Events}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{../res/comparison_plot.png}
  \caption{Decline in sentiment during a heat wave across several different neighborhoods, compared with other impacts on sentiment, such as major hurricanes, as well as the weekly variation in sentiment from the peak on Saturday to the low on Monday.}
  \label{fig:compare}
\end{figure}

@@

\subsection{Effects by Combined Statistical Area}

@@

\section{Discussion}
% Benefits of twitter data/overview of findings
We found that twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution.  We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between heat and worsened sentiment.   Moreover, we were able to show clear heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

@@

% NCC doesnt have a "conclusion" section, but this is intended as an overall summary paragraph
While climate change will have widespread and severe impacts on human well-being, it cannot be overemphasized just how highly unevenly these impacts will be distributed.  People with more money, access to aid and infrastructure, and who belong to ethnic groups in power are less affected by climate shocks and natural disasters and more able to adapt \cite{bullard2012wrong}.  While the physical health impacts of these shocks are more visible and easy to measure, the mental health impacts of climate change are also causing severe human suffering and there is no reason to believe they are not also highly unevenly distributed.  Thus, there are strong theoretical priors behind the hypothesis that low-income and marginalized people are more vulnerable to the mental health impacts of higher temperatures, even though previous work at coarse spatial scales has not found such an effect.  By using fine-scale twitter data, we showed that there are indeed stark difference in the mental health impacts of heat among neighborhoods in the United States.  These findings have significant implications for urban planning, climate and environmental justice, and mental health.

\section{Methods}
\subsection{Tweets \& Sentiment}
We used data from 243 million geo-located English-language tweets from the lower 48 states from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include retweets in our analysis.  Additionally, we excluded all tweets that contained weather-related terms, to ensure that the sentiment expressed in the tweets was reflective of a user's mental state, and not a commentary on the weather.

We assessed the sentiment expressed in the tweets using the VADER (Valence Aware Dictionary for sEntiment Reasoning) sentiment corpus \cite{gilbert_vader_2014}. VADER is a lexicon and rule-based sentiment analysis tool that is specifically attuned to sentiments expressed in social media because of several features, as:

@@

\subsection{Weather}
We used data on local weather conditions from the North American Land Data Assimilation System (NLDAS), a gridded product developed by several collaborative institutions, including NOAA, NASA, Princeton University, and the University of Washington.  This dataset is available at an hourly temporal resolution, and at 1/8th decimal degree spatial resolution, and integrates a large quantity of observation-based and modeled data  \cite{xia_continental-scale_2012}.  For the exact hour and location of each tweet, we extracted temperature, specific humidity, air pressure, total precipitation, shortwave radiation, and wind speed.

@@

\subsection{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where tweets were located.  Data was at the level of the census block group, the smallest unit for which the census releases public data.

@@

\subsection{Modeling}
We assessed how expressed sentiment is affected by wet bulb globe temperature using segmented regressions, controlling for precipitation, and shortwave solar radiation (sunshine), as well as the following spatio-temporal fixed effects: the day of the week, the time of day, the day of the year, the year, the month, and the county.

@@

\section{Acknowledgements}
%Official SESYNC wording
This work was supported by the National Socio-Environmental Synthesis Center (SESYNC) under funding received from the National Science Foundation DBI-1639145.  Additionally, cloud resources for this project were provided for the Microsoft Azure cloud by Microsoft's AI for Earth program, grant number .

This data came from Twitter via the University of Vermont’s (UVM) agreement with Twitter to access its streaming API - colloquially referred to as the Decahose.  The UVM special agreement with Twitter allows for access to this data for research and analysis purposes and we have complied with all the terms of service for Twitter and UVM. 
\printbibliography

\section{Author Contributions}
Author contributions: M.W.C. and P.B. designed research and modeling strategy; A.S. provided data; M.W.C and Z.L. prepared data; M.W.C. and J.O.G analyzed data; and M.W.C., J.O.G., J.L., and P.A.W. wrote the paper.

This project was co-lead by M.W.C. and J.O.G.

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
% \usepackage[
%   style=numeric,
%   citestyle=numeric-comp,
%   backend=biber,
%   doi=true,
%   natbib=true,
%   sorting=none
% ]{biblatex}
\usepackage{scalerel,xparse}

@@

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures}

@@

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in 250 million geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effects is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being with the impacts projected to worsen in the coming years \cite{pachauri2014climate}.  A large body of research has assessed the ongoing impacts on climate change on outcomes like agriculture, economic growth, and physical health (NEED CITATIONS). Academic and medical researchers have highlighted the relative paucity of empirical studies into climate impacts on mental health at multiple scales and recently called for more \cite{Berry2018Apr, hayes_climate_2018}.  Researchers have begun to answer these calls, finding associations between deteriorating environmental conditions and a sense of "ecological grief" from their experienced loss \cite{Cunsolo2018Apr}, disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, as well as a variety of mental health impacts associated with warmer temperatures \cite{baylis_weather_2018, Mullins2019Dec} (PALINKAS and WONG 2020).  MAYBE A SENTENCE HERE HIGHLIGHTING THE GAP IN KNOWLEDGE ABOUT DISPARITIES IN MENTAL HEALTH IMPACTS, VULNERABLE POPULATIONS, issues of scale for signal to noise, etc.

%Heat and mental health
%https://www.pnas.org/content/115/43/10953
Higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other work demonstrating similar relationships in other locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}.  Higher temperatures have also been correlated with increases in hospitalization events related to both general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb, Gupta1992Jun} as well as increased mortality among people with mental health disorders \cite{Hansen2008Oct}.

@@

Additionally, recent studies have found that nighttime temperatures affects sleep quality \cite{Obradovich2017May, Mullins2019Dec}.

% The pathways
Higher temperature associated with climate change could influence mental health both directly and indirectly by exposing people to trauma (direct) (CITATIONS) or by affecting (1) physical health and (2)community well-being (indirect) \cite{BerryETAL2010}. Direct pathways can include extreme heat events or increasing temperatures associated with loss of workplace productivity from decreases in occupational health or increases in workplace injury \cite{kjellstrom_heat_2016, KjellstromETAL2009, kjellstrom_impact_2016} which can result in loss of income, all of which can exacerbate deteriorating mental health and lead to increased aggression, hospital admissions and suicide rates \cite{BerryETAL2010,CheatwoodDerral1995,CohnETAL2004,BouchamaETAL2007}. Heat exposures in working environments could also reduce people's capacity to deal with physical and mental task, and increase the risk of accidents, because of the excessive core body temperature and dehydration \cite{KjellstromETAL2009}. The loss of work capacity would in turn result in loss of income, while the low income could also cause mental health problems \cite{Katz1997}.  As for the indirect pathway, first, because of the reciprocal causation relationship between physical health problems and mental health problems \cite{MillerETAL2009,PrinceETAL2007} heat exposures associated with climate change along with other climate events and indirect health risks threatening physical health will directly influence mental health \cite{Berry2007,WHO2007}. Second, disordered temperature associated with climate change may destroy the economic in agricultural-production-dependent communities. For example, extreme heat reduces the work capacity of laborers in farm fields \cite{KjellstromETAL2009}, which further destroys agricultural-supported industries in local area\cite{BerryETAL2008}. The following economic pressures would undermine social capital and then lead to mental health problems.

@@

\section*{Results}

\subsection*{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  Controlling for a variety of fixed effects, we found that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree Celsius (a dry bulb temperature typically around 12\textdegree C/54\textdegree F).  The greatest declines in sentiment are observed between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We ran models using multiple sentiment indicators, and found similar results no matter which sentiment metric was used (See Supplement).

@@

\subsection*{Comparison With Other Events}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
  \caption{Decline in sentiment during a heat wave across several different neighborhoods, compared with other impacts on sentiment, such as major hurricanes, as well as the weekly variation in sentiment from the peak on Saturday to the low on Monday.}
  \label{fig:compare}
\end{figure}

@@

\subsection*{Effects by Combined Statistical Area}

@@

\section*{Discussion}
% Benefits of twitter data/overview of findings
We found that twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution.  We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between heat and worsened sentiment.   Moreover, we were able to show clear heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

@@

% NCC doesnt have a "conclusion" section*, but this is intended as an overall summary paragraph
While climate change will have widespread and severe impacts on human well-being, it cannot be overemphasized just how highly unevenly these impacts will be distributed.  People with more money, access to aid and infrastructure, and who belong to ethnic groups in power are less affected by climate shocks and natural disasters and more able to adapt \cite{bullard2012wrong}.  While the physical health impacts of these shocks are more visible and easy to measure, the mental health impacts of climate change are also causing severe human suffering and there is no reason to believe they are not also highly unevenly distributed.  Thus, there are strong theoretical priors behind the hypothesis that low-income and marginalized people are more vulnerable to the mental health impacts of higher temperatures, even though previous work at coarse spatial scales has not found such an effect.  By using fine-scale twitter data, we showed that there are indeed stark difference in the mental health impacts of heat among neighborhoods in the United States.  These findings have significant implications for urban planning, climate and environmental justice, and mental health.

\section*{Methods}
\subsection*{Tweets \& Sentiment}
We used data from 243 million geo-located English-language tweets from the lower 48 states from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include retweets in our analysis.  Additionally, we excluded all tweets that contained weather-related terms, to ensure that the sentiment expressed in the tweets was reflective of a user's mental state, and not a commentary on the weather.

We assessed the sentiment expressed in the tweets using the VADER (Valence Aware Dictionary for sentiment Reasoning) sentiment corpus \cite{gilbert_vader_2014}. VADER is a lexicon and rule-based sentiment analysis tool that is specifically attuned to sentiments expressed in social media because of several features, as:

@@

\subsection*{Weather}
We used data on local weather conditions from the North American Land Data Assimilation System (NLDAS), a gridded product developed by several collaborative institutions, including NOAA, NASA, Princeton University, and the University of Washington.  This dataset is available at an hourly temporal resolution, and at 1/8th decimal degree spatial resolution, and integrates a large quantity of observation-based and modeled data  \cite{xia_continental-scale_2012}.  For the exact hour and location of each tweet, we extracted temperature, specific humidity, air pressure, total precipitation, shortwave radiation, and wind speed.

@@

\subsection*{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where tweets were located.  Data was at the level of the census block group, the smallest unit for which the census releases public data.

@@

\subsection*{Modeling}
We assessed how expressed sentiment is affected by wet bulb globe temperature using segmented regressions, controlling for precipitation, and shortwave solar radiation (sunshine), as well as the following spatio-temporal fixed effects: the day of the week, the time of day, the day of the year, the year, the month, and the county.

@@

\section*{Acknowledgements}
%Official SESYNC wording
This work was supported by the National Socio-Environmental Synthesis Center (SESYNC) under funding received from the National Science Foundation DBI-1639145.  Additionally, cloud resources for this project were provided for the Microsoft Azure cloud by Microsoft's AI for Earth program, grant number .

This data came from Twitter via the University of Vermont’s (UVM) agreement with Twitter to access its streaming API - colloquially referred to as the Decahose.  The UVM special agreement with Twitter allows for access to this data for research and analysis purposes and we have complied with all the terms of service for Twitter and UVM.

%\printbibliography
\bibliography{TextDataClimateShocks}

\section*{Author Contributions}
Author contributions: M.W.C., J.O.G. and P.B. designed the research and modeling strategy; A.S. provided data; M.W.C and Z.L. prepared data; M.W.C. and J.O.G analyzed data; and M.W.C., J.O.G., J.L., and P.A.W. wrote the paper.

\noindent This project was co-lead by M.W.C. and J.O.G.
